---
title: "Environmental Factor's effect on Redlined Communities"
author: "Annie Adams"
date: "Dec 10, 2023"
output:
  html_document:
    print_df: paged
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Background

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.



## Overview
This repo will explore both the relationships between census groups/HOLC groups with environmental factors. We will first look at the relationship between census groups and environmental factors. Using historical redlining data, we will then explore similar environmental factors as we did with census groups, but with HOLC grades instead. We will see if we can find any noticeable differences amongst different HOLC grades to see if redlining still plays an effect in these neighborhoods. We will explore these differences through both visual and numerical analysis. 



## Datasets:

#### EJScreen

We will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.

#### Biodiversity observation data

The [Global Biodiversity Information Facility](https://www.gbif.org/) is the largest aggregator
of biodiversity observations in the world. Observations typically
include a location and date that a species was observed.We will be
working observations of birds from 2021 onward. 

#### Inequality Maps

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^4]

[^4]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

## Analysis

### Load Necessary packages and data

```{r }
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(geodata)
library(tidyverse)
library(here)
library(sf)
library(gt)
library(kableExtra)

```

Read in EJScreen data and filter to Los Angeles County

```{r echo=TRUE, results='hide'}
ejscreen <- st_read("../data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/") 

la_data <- ejscreen %>% filter(CNTY_NAME == "Los Angeles County") # filter data to LA
```

Import redlining information for Los Angeles.

```{r echo=TRUE, results='hide'}

LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid()
```

Read in bird data from Global Biodiversity Information Facility.

```{r echo=TRUE, results='hide'}
bird <- st_read("../data/gbif-birds-LA") 
```

### Create wastewater map across census groups

Different environmental factors affect different censuses in disproportionate ways. To visualize this, we will map wastewater discharge.To see the upper values for this variables, we will select census block groups that are greater than the 95th percentile.

```{r }
wastewater_95_perc <- la_data %>% filter(P_PWDIS > 95)
#la_centroids <- st_centroid(la_data)
la_over95_centroids <- st_centroid(wastewater_95_perc)

wastewater_map <-ggplot()+ 
  geom_sf(data = la_data, aes(fill =P_PWDIS)  )+
  geom_sf(data = la_over95_centroids, aes(fill = P_PWDIS))+
  labs(fill='Percentile Wastewater', x = "Longitude", y = "Latitude")+
  scale_fill_gradient(low = "#56B1F7", high = "#132B43")+
  theme_minimal()

  
  
wastewater_map

```

This map allows us to visualize which areas of LA county have a higher
percent of wastewater. Gray areas are areas in which no wastewater
percentile was reported.



#### Low income percentage in census groups

Let's now look more closely at other relationships amongst block groups
and environmental factors. We will start by finding the percent of
census block groups where less than 5% of the population is considered
low income.

```{r } #P_LOWINCPCT}
less_5_inc <- la_data %>% filter(LOWINCPCT <.05)
perc_below5_income <- round(length(less_5_inc$LOWINCPCT)/ length(la_data$LOWINCPCT), 3) * 100
print(paste0("The percent of census block groups where less than 5% of the population is considered low income is ", perc_below5_income, "%."))
```


6.1% is a pretty small number when discussing percentages not considered
low income. This means that in approximately 94% of census groups in LA,
the population considered low income is greater than 5%.



#### Census groups and air quality experiences

We will continue to look at percentages amongst block groups for some
different environmental factors, specifically what percent of census
block groups that are both above the 80th percentile for PM 2.5 as well
as the 80th percentile for the Superfund proximity ( a superfund is an
abandoned hazardous waste site). These combination of different
variables allow us to dive specifically into how variables related to
air quality effect different census groups.

```{r } # P_PM25}
above_80_pm25_pnpl <- la_data %>% filter(P_PM25 >80, P_PNPL>80)
perc_above80_pm25_pnpl <- round(length(above_80_pm25_pnpl$P_PM25)/ length(la_data$P_PM25), 3) * 100
print(paste0("The percent of census block groups above the 80th percentile for both PM 2.5 and Superfund proximity is ", perc_above80_pm25_pnpl, "%."))

```


#### Mapping Redlined Neighborhoods by HOLC grades

We will now utilize the LA Redlining data we loaded in. Let's start
with some exploratory analysis to get a better idea of what our data
looks like. We will simply create a map of the different HOLC grades in
LA County.

```{r }
holc_colors <- c(
  "A" = "#9CAF88",  
  "B" = "#007BB8",  
  "C" = "#FBE790",  
  "D" = "#F2B8C6"   
)


redlining_map <-ggplot()+ 
  #geom_sf(data = la_data, fill = NA)+
  geom_sf(data = LA_redlining, aes(fill = grade))+
   scale_fill_manual(values = holc_colors)+
  labs(fill='HOLC Grade', x = "Longitude", y = "Latitude", title = "Historical redlining boundaries colored by HOLC grade") +
  theme_minimal()
redlining_map

ggsave(filename = "../Results/HOLC_grade_map.png", plot = last_plot())
```

#### Census groups within HOLC grades

Now that we have visualized the geographical relationships of different
HOLC grades in LA county, lets look a bit more into the specifics of census groups within the different HOLC grades. We will start by
determining how many different census groups are within the four
different HOLC grades.

```{r eval=FALSE}
# Check crs for both datasets
st_crs(LA_redlining) == st_crs(la_data)
la_data <- st_transform(la_data, crs = 4326) # change la data to have crs 4326


cen_holc<- st_join(x = la_data, y = LA_redlining, join = st_within, left = FALSE) # join la data and redlining data using st_within to find where areas with HOLC grades
census_group_HOLC <-nrow(cen_holc) # find number of census block groups that fall within HOLC grades

print(paste0("There are ",census_group_HOLC, " census block groups that fall within HOLC grade areas." ))
```


#### Environmental experiences within HOLC grades

We looked into how different environmental factors impact census groups,
but now lets do the same with HOLC grades instead of census groups.
Doing so will allow us to see if different environmental factors
disproportionately affect different HOLC grades. We will look at four
different environmental factors: average percent low income, average
percent of PM 2.5 present, average percentile for low life expectancy,
and average percentile for air toxics cancer risk.

In order to do this, we need to start by joining our redlined data with
the EJ screen data from LA.

```{r }
la_data <- st_transform(la_data, crs = 4326) # change la data to have crs 4326
redline_ej_join<- st_join(LA_redlining, la_data) # join redline and la data 
```

```{r}


redline_ej_join_test <- redline_ej_join %>%
  group_by(grade) %>%
  summarise( #create average variable for each of the four catergories we are looking into, drop NAs and calculate mean
    "Average low income percent" = mean(LOWINCPCT, na.rm = TRUE), 
    "Average percent PM 25" = mean(P_PM25, na.rm = TRUE),
    "Average percentile for low life expectancy"  = mean(LIFEEXPPCT, na.rm = TRUE),
    "Average percentile for air toxics cancer risk" = mean(P_CANCER, na.rm = TRUE)
  ) %>% 
  kbl(linesep = "", booktabs = TRUE) %>% #make table pretty
  kable_styling(bootstrap_options = c("striped", "condensed"),
                latex_options = c("striped", "hold_position"),full_width = FALSE)

redline_ej_join_test

```

#### What do these results mean?

The values calculated above prove that lower grades experience the
effect of the environment at increased rates.The average percent of low
income is nearly 25% higher in neighborhoods graded as D over
neighborhoods graded as A. Similarly PM 2.5 increases as HOLC grade
decreases. C and D holc grades have a higher average percentile for low
life expectancy, as well as being more prone to cancer risk exposure.



#### Biodiversity and HOLC Grades

The variables we have looked into thus far have had a lot to do with air
quality/life expectancy. Let's look into the more recreational side of
the environment to see if differences can be seen there too. One
specific outdoor recreation that the EJ screen data reports on is bird
watching. Reports of bird observations help obtain a better
understanding of the biodiversity that exists within our environment.
Based on other environmental factors that do not see equal rates amongst
HOLC grades, we might not expect bird watching/ observations to be the
same amongst HOLC grades either. Let's find out by finding the percent
of bird observations within each HOLC grade.


To get a better idea about bird watching amongst different communities,
check out this
[article](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw)
by Ellis-Soto and co-authors. They found that redlined neighborhoods
remain the most undersampled areasfor bird observations across 195 US
cities. This gap is highly concerning, because conservation decisions
are made based on these data.[^5]

[^5]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.


We have data on both bird observations and redlining categories. Let's
join these datasets together, first making sure their crs' match. Then,
we will create find the percentage of bird observations within each HOLC
grade for the year 2022.

```{r }
st_crs(LA_redlining) == st_crs(bird) # check to make sure CRS match
bird_redlining<- st_join(y = LA_redlining, x = bird, join = st_intersects, left = FALSE) # join bird data and HOLc grade data by intersection
bird_redlining_2022 <- bird_redlining %>% filter(year == 2022) # filter for 2022

bird_redlining_2022<- bird_redlining_2022 %>%
  group_by(grade) %>%                  
  summarise(count = n()) %>%        #count number of obs within each holc grade      
  mutate(percentage = (count / sum(count))*100 ) #compare specific HOLC grade to total obs and make into a percent
```


#### Visualize Biodiversity amongst HOLC grades
Lets plot the results from our table above to get a better idea about
how bird observations differe across HOLC grades.

```{r}


holc_colors <- c(
  "A" = "#9CAF88",  
  "B" = "#007BB8",  
  "C" = "#FBE790",  
  "D" = "#F2B8C6"   
)

ggplot(bird_redlining_2022, aes(x = grade, y = percentage, fill = grade))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = holc_colors)+
  labs( x = "HOLC grade", y = "percentage of bird observations", fill = "HOLC grade", title = "Percentage of bird observations across HOLC grades")+
  theme_minimal()
```




#### Is this what we expected ??

These results were very surprising to me. After reading *Bird
Biodiversity Reports Reflect Citie's Redlined Past,* I expected HOLC
grades A and B to have a far greater percentage of bird observations
then grades C and D. The article mentions how crowd-sourced bird
watching data comes from White neighborhoods that are likely to reside
in A or B HOLC grades. The article puts forth a study that looks into
bird observations among the four different HOLC grades from 1932 to
2022. Their study finds a larger percentage of bird observations to be
from HOLC grade A than any other HOLC grade, in Los Angeles. This
contradicts the plot above significantly. The main difference in this
plot above and the one put forth in the article is the time period
observed. This bar plot looks only at 2022, while the other spans nearly
100 years. Perhaps there has been lasting effects of COVID 19 that have
caused community members in lower HOLC grades to spend more time outside
looking at birds.\*\*
